They're Clutching up! Team Momentum in Round-Based Esports

### 2022-08-12 paper/paper.qmd
do a test render
paper

---
title: "They're Clutching up! Team Momentum in Round-Based Esports"
format:
  arxiv-pdf:
    keep-tex: true  
  arxiv-html: default
author:
  - name: Tony ElHabr
    affiliations:
      - name: Georgia Institute of Technology
    email: anthonyelhabr@gmail.com
    url: https://tonyelhabr.rbind.io
abstract: |
  My research investigates patterns in round win percentages in professional matches (series) of two popular first-person shooter games, Call of Duty (Search and Destroy mode) and Valorant.

  First, I find evidence in Call of Duty (COD) defying the naive hypothesis that a series represents a sequence of independent events (rounds), with each team having a constant 50% probability of winning a given round.

  Second, I examine post-streak round win probability. For COD, in which teams play first to 6 round wins (maximum 11 rounds), I find that teams perform significantly worse than expected after streaks of 2, 3, and 4 wins when series end up going to 9, 10, or 11 rounds, even after accounting for the "hot-hand" phenomenon. In Valorant, which requires one side to win 13 rounds for the series victory, there is evidence of performing better than expected after streaks of up to 8 wins, which is perhaps not shocking given the economy aspect of the game.

  Third, I compare win percentages in round one versus all other rounds, hypothesizing that there may be some advantage on either side when there is no prior information about how the opponent intends to play a given map in either game. I find one COD map for which there seems to be a significant defensive advantage in round one, but none otherwise.

  Finally, I evaluate behavior when teams have two rounds left to win the series, observing a peak in COD offensive win percentages in the 4-4 state, and no such oddity in Valorant.
keywords: 
  - esports
bibliography: bibliography.bib  
---

Sit nunc at convallis fringilla semper! Penatibus vivamus eget malesuada cursus fames magnis potenti. Venenatis ligula enim conubia laoreet orci class ligula? Massa sem sed enim risus ut mattis inceptos nisl elementum mattis. Nostra ultricies habitant donec vitae luctus proin gravida placerat ac libero imperdiet. Auctor habitasse faucibus ultricies purus eget sociis ultrices habitasse nec ad aptent lectus rutrum proin potenti montes sodales posuere convallis tempor erat egestas magna lectus sociis mollis purus.

Dolor turpis euismod himenaeos interdum felis dictum tempus euismod tortor aliquam? Eget interdum vehicula laoreet quam. Mollis justo cursus ad blandit feugiat pulvinar sem sodales bibendum. Leo leo magna pulvinar metus lacinia nam fringilla maecenas duis dis suscipit aenean natoque sem metus quam risus sagittis convallis primis sociis id dictumst.

# Introduction

Amet ultricies dignissim varius potenti suscipit justo phasellus dui ullamcorper luctus suscipit vulputate. Sagittis enim volutpat ultrices diam curabitur potenti platea euismod vel libero primis feugiat. Nibh potenti habitant ultricies dictum sociis class at vehicula eu vestibulum. Curabitur iaculis.

Consectetur est eleifend turpis morbi magna porttitor bibendum ut blandit aliquet lacus. Nostra iaculis torquent parturient id fermentum montes at himenaeos. Magna congue blandit fames egestas at inceptos? Id maecenas blandit maecenas accumsan ridiculus mollis nullam justo pulvinar elementum inceptos habitant risus.

{{< pagebreak >}}

>>>>

---
title: "They're Clutching up! Team Momentum in Round-Based Esports"
format:
  arxiv-pdf:
    keep-tex: true  
  arxiv-html: default
author:
  - name: Tony ElHabr
    affiliations:
      - name: Georgia Institute of Technology
    email: anthonyelhabr@gmail.com
    url: https://tonyelhabr.rbind.io
abstract: |
  My research investigates patterns in round win percentages in professional Search and Destroy (SnD) matches of the popular first-person shooter game Call of Duty (CoD).

  First, I find evidence in CoD defying the naive hypothesis that a series represents a sequence of independent events (rounds), with each team having a constant 50% probability of winning a given round.

  Second, I examine post-streak round win probability. I find that teams perform significantly worse than expected after streaks of 2, 3, and 4 wins when series end up going to 9, 10, or 11 (maximum) rounds, even after accounting for the "hot-hand" phenomenon.

  Third, I compare win percentages in round one versus all other rounds, hypothesizing that there may be some advantage on either side when there is no prior information about how the opponent intends to play a given map in either game. I find only one instance for which there seems to be a significant defensive advantage in round one.

  Finally, I evaluate behavior when teams have two rounds left to win the series, observing a peak in COD offensive win percentages in the 4-4 state, and no such oddity in Valorant.
bibliography: references.bib  
---

# 1. Introduction

## 1.1 Description of Call of Duty Search and Destroy

Call of Duty (CoD), first released in 2003, is one of the most popular first-person shooter (FPS) video game franchises of all-time. Tthe most popular mode in the competitive scene is "Search and Destroy" (SnD), which bears resemblance to "Bomb Defusal" in Counter-Strike and "Plant/Defuse" in Valorant, two other FPS games played in professional leagues. SnD is one-sided game mode in which one team, the offensive side, tries to destroy one of two designated bomb sites on the map.[^1]

[^1]: as

In professional CoD, a team must win six rounds of SnD to win the match.[^2] A round can end in one of four ways:

[^2]: A maximum of 11 even rounds can be played. There is no "sudden death" or "win by two" rule like there are for SnD equivalent in professional Counter-Strike and Valorant matches.

1.  One team eliminates all members of the other team prior to a bomb plant. (Eliminating team wins.)
2.  The offensive team eliminates all members of the defensive team after a bomb plant.[^3] (Offense wins.)
3.  The defensive team defuses the bomb after a bomb plant.[^4] (Defense wins.)
4.  The offensive team does not make a plant by the time the round timer ends. (Defense wins.)

[^3]:
    -   The bomb can be picked up by any member of the offensive team.
    -   The bomb carrier is not obstructed at all by carrying the bomb (i.e. movement is the same, weapon usage is the same).
    -   The defense does not get any visual indication for who is carrying the bomb.
    -   A bomb plant takes five seconds. The timer resets if the player stops planting site prior to completing it.
    -   A bomb defuse takes seven seconds. The timer resets if the player "drops" the bomb.
    -   The bomb takes 45 seconds to defuse after being planted.

[^4]: Often the defensive team will try to eliminate all team members prior to making the defuse, but in some cases, they may try to "ninja" defuse.

Teams take turns playing offense and defense every round.

## 1.2 Data

CoD has roughly gone through three eras of professional gaming: (1) Major League Gaming (MLG) tournaments prior to 2016; (2) the CoD World League (CWL), initiated in 2016; and (3) the 12-franchise CoD League (CDL), running since 2020 and completing three year-long "seasons" completed as of August 2022. The data set consists of all SnD matches played in tournaments and qualifiers during the CDL era. Data was collected in spreadsheets by community member "IOUTurtle".[^5]

[^5]: Data: https://linktr.ee/CDLArchive. Author: https://twitter.com/IOUTurtle

I adopt the terminology "series" to refer to what CoD SnD players typically call a "match", so as to emulate the terminology of playoff series in professional leagues like the National Basketball Association, National Hockey League, and Major League Baseball. A "game" or a "match" in such leagues is analogous to a "round" of CoD SnD.

CoD is fairly unique compared to other esports in that it runs on an annual lifecycle, where a new game is published every year under the same title. Each new game bears resemblance to past ones, often introducing relatively small variations ("improvements") to graphics, game modes, and other facets of gameplay. The data set denotes which CoD game a given CoD SnD series was played,

## 1.3 My contribution

# 2 Literature review

Streaks and momentum is one of the most frequently published topics in sports analytics. We often use these interchangeably, but as [@steeger2021] note, momentum implies dependence between events, whereas streaking does not. When considering

Despite the plethora of existing research on momentum in sports, its yet to be investigated heavily in esports. [@article-journal]
